---
title: "R Notebook"
output: html_notebook
---

```{r load modules}
library(tidyverse)
```

-- Read in the methylation data as well as the meta data --
```{r}
# Cells with pseudotime (and other info) -> meta data
cells_ptime <- read_tsv("/mnt/raid/leoni/sc_data_lukas/heibox_data/21-12-01_scNMT-cell-metadata.tsv") %>%
  filter(!is.na(ptime))


Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 100)

# Total sites per VMR
VMR_met_sites <- read_csv("/mnt/raid/leoni/sc_data_lukas/heibox_data/meth_coverage_vr.csv.gz") %>%
  pivot_longer(!...1, names_to = "VMR", values_to = "total_sites") %>%
  rename(sample = ...1)

# Percentage of methylated VMRs
VMR_met_per <- read_csv("/mnt/raid/leoni/sc_data_lukas/heibox_data/meth_mfrac_vr.csv.gz") %>%
  pivot_longer(!...1, names_to = "VMR", values_to = "met_percentage") %>%
  filter(!is.na(met_percentage)) %>%
  rename(sample = ...1) 
```
-- Join methylation and metadata together --
```{r}
# Met VMRs and ptime for cells in one
VMR_met_ptime <- VMR_met_per %>%
  left_join(VMR_met_sites, by = c("sample", "VMR")) %>%
  mutate(met_sites = (met_percentage*total_sites)) %>%
  inner_join(cells_ptime %>% select(c("sample", "ptime", "celltype"))) %>%
  rename(pseudotime = ptime , methylated_sites = met_sites)
```

-- UMAPs --
```{r}
ggplot(cells_ptime, aes(x = UMAP1, y = UMAP2, col = celltype)) +
  geom_point(size = .5) +
  coord_fixed()+
  labs(x = "UMAP1",
       y = "UMAP2",
       color = "Celltype", 
       tag = "c") +
  theme_classic()

ggplot(cells_ptime, aes(x = UMAP1, y = UMAP2, col = ptime)) +
  geom_point(size = .5) +
  coord_fixed()+
  labs(x = "UMAP1",
       y = "UMAP2",
       color = "Pseudotime", 
       tag = "d") +
  theme_classic()
```


-- Calculate the average methylation of a VMR before and after a point in pseudotime and the difference between those to find out at which point in pseudotime the VMR goes from mostly (un)methylated to the opposite --
```{r}

# select a VMR of interest
VMR_met_ptime_one_VMR <- VMR_met_ptime %>%
  filter(VMR == "X_99138536_99154786") 

# Plot met before, after and diff 
VMR_met_ptime_one_VMR %>%
arrange( pseudotime ) %>%
mutate(
  avg_meth_before = cumsum(methylated_sites) / cumsum(total_sites),
  avg_meth_after = (sum(methylated_sites)-cumsum(methylated_sites)) / (sum(total_sites)-cumsum(total_sites))
) %>%
  mutate( avg_meth_diff = ifelse( 
    cumsum(methylated_sites) > 15 & sum(methylated_sites) - cumsum(methylated_sites)>15,
    abs(avg_meth_after - avg_meth_before), NA ) )%>%
ggplot() + geom_jitter( aes( x=pseudotime, y=met_percentage*100 ), width=0, height=.03 ) +
  geom_line(aes(x=pseudotime, y=avg_meth_before*100, col="Average methylation before"), size=1 ) +
  geom_line(aes(x=pseudotime, y=avg_meth_after*100, col="Average methylation after"), size=1 ) +
  geom_line(aes(x=pseudotime, y=avg_meth_diff*100, col = "Difference"), size=1 ) +
  labs(y = "Methylated sites [%]",
       x = "Pseudotime" ,
       color = "",
       tag = "e") +
  theme_classic()

```


-- Find the Inflection Points for all VMRs --
```{r Inflection point}

find_infl_VMR <- function( a, id ) {
  a %>%
    arrange( pseudotime ) %>%
    mutate( 
      avg_meth_before = cumsum(methylated_sites) / cumsum(total_sites),
      avg_meth_after = (sum(methylated_sites)-cumsum(methylated_sites)) / (sum(total_sites)-cumsum(total_sites))
  ) %>%
  mutate( avg_meth_diff = ifelse( 
    cumsum(methylated_sites) > 15 & sum(methylated_sites) - cumsum(methylated_sites)>15,
    avg_meth_after - avg_meth_before, 0 ) ) %>%
  summarise(
    gene_id = id,
    infl_cell = which.max( abs( avg_meth_diff ) ),
    infl_point = pseudotime[ infl_cell ],
    meth_diff = avg_meth_diff[ infl_cell ],
    total_sites = sum(total_sites)
  )
} 


# Find the Inflection point for all (takes ~20 min)
VMR_met_ptime %>% 
  filter_at(vars(pseudotime,total_sites, methylated_sites),all_vars(!is.na(.))) %>%
  group_by( VMR ) %>%
  group_map( find_infl_VMR )  %>%
  bind_rows() %>%
  unnest(gene_id)-> infl_point_all_VMR

# Visualization
ggplot(infl_point_all_VMR %>% filter(meth_diff != 0), aes(x = infl_point, y = meth_diff)) + 
  geom_point(size = .5) +
  labs(y = "Methylation difference",
       x = "Pseudotime",
       tag = "d") +
  theme_classic()
  
```

 -- Smoothing --
```{r}
VMR_met_ptime %>% group_by( VMR ) %>% summarise() -> all_VMR

VMR_met_ptime %>% select(- celltype, -met_percentage, -sample) %>%
  group_by(VMR) -> VMR_met_ptime_grouped

VMR_met_ptime_grouped %>%
  group_split()-> VMR_met_ptime_split

pt_grid_VMR <- seq( min(VMR_met_ptime$pseudotime, na.rm = TRUE), max(VMR_met_ptime$pseudotime, na.rm = TRUE), length.out = 100)

smooth_met_neuro <- function( VMR_in ){
  
  
  VMR_met_ptime_split[[which(group_keys(VMR_met_ptime_grouped) == VMR_in)]] -> b
  
  tib <- tibble(met_sites = b$methylated_sites,
                total_sites = b$total_sites,
                pseudotime = b$pseudotime)
  
  
  
  smooth_value <- sapply(pt_grid_VMR, function(pt) {
    weight <- kernel( pt - tib$pseudotime ) / .1 
    sum(tib$met_sites*weight) / sum(tib$total_sites*weight)
  })
  
  names(smooth_value) <- round( pt_grid_VMR, 2 )
  smooth_value
}

# ~15 min
sapply(all_VMR$VMR, smooth_met_neuro) -> smoothed_neuro

```

-- Smoothing example one VMR --
```{r}
smooth_met_example_VMR <- smooth_met_neuro( "X_99138536_99154786" ) %>%
  enframe(name = "pt", value = "smoothed_value") %>%
  mutate(pt = as.numeric(pt))

VMR_met_ptime %>%
  filter(VMR == "X_99138536_99154786") %>%
ggplot()+
  geom_point(aes(x = pseudotime, y = met_percentage*100)) +
  geom_line(data = smooth_met_example_VMR, aes(x = pt, y = smoothed_value*100, group = 1, col = "red")) +
  labs(y = "Methylation [%]",
       x = "Pseudotime",
       tag = "b") +
  theme_classic() +
  theme(legend.position = "none")
```


-- Heatmap --
```{r}

set.seed(12345)
smoothed_neuro[,sample.int( ncol(smoothed_neuro), 3000 )]*100 -> neuro_matplot_3000
t(neuro_matplot_3000) -> neuro_heatmap_3000

matplot( x=pt_grid_VMR,y=neuro_matplot_3000, type = "l", col= alpha("black", 0.1), lty = "solid", xlab = "Pseudotime", ylab = "Methylation [%]")
mtext("d", at=-.8, line=2)


fields::image.plot( x=as.numeric(colnames(neuro_heatmap_3000)), z=( t(neuro_heatmap_3000[order(neuro_heatmap_3000[,23] - neuro_heatmap_3000[,28] ),]) ), ylim=c(-.1,1), col = viridisLite::cividis(100), zlim=c(0,100), xlab="Pseudotime", ylab="VMRs", yaxt="n" )
segments( unique(VMR_met_ptime$pseudotime), -.8, unique(VMR_met_ptime$pseudotime), -.03, col=alpha("black",.3) )
mtext("f", at=-.8, line=2)

```

